Subsidiary Recursion in Coq

### 2022-01-08 paper/paper.tex
working on paper
Outline is there now, working on describing the interface we provide

\author{Aaron Stump}{Computer Science Dept., The University of Iowa, USA \and \url{http://www.cs.uiowa.edu/~astump/}}{aaron-stump@uiowa.edu}{http://orcid.org/0000-0002-9720-0003}{ }%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\authorrunning{A. Stump} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Aaron Stump} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\keywords{strong functional programming, recursion schemes, impredicativity} %TODO mandatory; please add comma-separated list of keywords

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
\end{abstract}

\section{Start}

>>>>

\author{Aaron Stump}{Computer Science Dept., The University of Iowa, USA \and \url{http://www.cs.uiowa.edu/~astump/}}{aaron-stump@uiowa.edu}{http://orcid.org/0000-0002-9720-0003}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Alex Hubers}{Computer Science, The University of Iowa, USA}{alexander-hubers@uiowa.edu}{}{}

\author{Christopher Jenkins}{Computer Science, The University of Iowa, USA}{alexander-hubers@uiowa.edu}{http://orcid.org/
0000-0002-5434-5018}{}

\author{Benjamin Delaware}{Computer Science, Purdue University, USA \and \url{https://www.cs.purdue.edu/homes/bendy/}}{bendy@purdue.edu}{}{}

\authorrunning{A. Stump, A. Hubers, C. Jenkins, and B. Delaware} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Aaron Stump, Alex Hubers, Christopher Jenkins, and Benjamin Delaware} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\keywords{strong functional programming, recursion schemes, positive-recursive types, impredicativity} %TODO mandatory; please add comma-separated list of keywords

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
  This paper describes a functor-generic derivation of a recursion
  scheme supporting the initiation of subsidiary recursions, in Coq.
  The approach supports making outer recursive calls on the results of
  subsidiary recursions.  The derivation utilizes a novel (necessarily
  weakened) form of positive-recursive types in Coq.  Subsidiary
  induction is also supported.  Using this derivation, several
  examples are demonstrated.
\end{abstract}

\section{Introduction: subsidiary recursion}

Central to interactive theorem provers like Coq, Agda, Isabelle/HOL,
Lean and others are terminating recursive functions over user-declared
inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
usually enforced by a syntactic check for structural decrease.  This
structural termination is sufficient for many basic functions.  For
example, the well-known \texttt{span} function from Haskell's standard
library (\verb|Data.List|) takes a list and returns a pair of the
maximal prefix satisfying a given predicate \verb|p|, and the
remaining suffix:
\begin{verbatim}
span :: (a -> Bool) -> [a] -> ([a],[a])
span _ []    =  ([], [])
span p (x:xs) = if p x
                then let (ys,zs) = span p xs in (x:ys,zs)
                else ([],x:xs)
\end{verbatim}
\noindent The sole recursive call is \verb|span p xs|, and it occurs
in a clause where the input list is of the form \verb|x:xs|.  The
input to the recursive call is a subdatum of the input for that
clause, and hence this definition is structurally decreasing.  In the
appropriate syntax, it can be accepted without additional effort by
all the mentioned provers.

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further recursion.  Let us see a simple example.  The
function \verb|wordsBy| (\verb|Data.List.Extra|) breaks a list into
the list of its maximal sublists whose elements do not satisfy a
predicate \verb|p|.  For example, \verb|wordsBy isSpace " good day "|
returns \verb|["good","day"]|; so \verb|wordsBy isSpace| has the same
behavior as \verb|words| (\verb|Data.List|).  Code is in
Figure~\ref{fig:wordsBy}. The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination.

\begin{figure}
\begin{verbatim}
wordsBy :: (a -> Bool) -> [a] -> [[a]]
wordsBy p [] = []
wordsBy p (hd:tl) =
  if p hd
  then wordsBy p tl 
  else let (w,z) = span (not . p) tl in
        (hd:w) : wordsBy p z
\end{verbatim}
\caption{Haskell code for \texttt{wordsBy}, demonstrating subsidiary recursion}
\label{fig:wordsBy}
\end{figure}

\subsection{Summary of results}

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We should emphasize that
this is a derivation of this recursion scheme within the type theory
of Coq.  (So no axioms or other modifications to Coq of any kind are
required.) Based on this derivation, we present several example
functions like \verb|wordsBy|, and prove theorems about them.  For
example, we prove the expected property that the sublists returned by
\verb|wordsBy| contain no elements satisfying the predicate \verb|p|.
For another, we give a definition of run-length encoding as a
subsidiary recursion using \verb|span|, and prove that encoding and
then decoding returns the original list.  The approach applies to the
standard datatypes in the Coq library, and does not require switching
libraries or datatype definitions.

An important technical novelty of our approach is a derivation of a
weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
restrict datatypes $D$ to be strictly positive: in the type for any
constructor of $D$, $D$ cannot occur to the left of any arrows.  Our
derivation needs to use positive-recursive types, where $D$ may occur
to the left only of an even number of arrows.  Full positive-recursive
types are incompatible with Coq's type theory.  But we present a way
to derive a weakened form that is sufficient for our examples
(Section~\ref{sec:mu}).  The weakening is to require just a retraction
between recursive type $\mu$ and its one-step unfolding $F\ \mu$,
instead of an isomorphism.  Hence, we dub these
\textbf{retractive-positive} recursive types.

We begin by summarizing the interface our derivation provides for
subsidiary recursion (Section~\ref{sec:interface}), and then see
examples (Section~\ref{sec:examples}).  The interface for subsidiary
induction is covered next (Section~\ref{sec:interfacei}), and example
proofs using it (Section~\ref{sec:examplesi}).  We conclude with a
walk-through of the derivation of that first interface (for subsidiary
recursion), in Section~\ref{sec:deriv}.  Related work is discussed in
Section~\ref{sec:related}.

All presented derivations have been checked with Coq version 8.13.2.
The code may be found as release \verb|itp-2022| (dated prior to the
ITP 2022 deadline) at
\url{https://github.com/astump/coq-subsidiary}.

\section{Interface}
\label{sec:interface}

\begin{figure}
  \begin{verbatim}
Inductive ListF(X : Set) : Set :=
| Nil : ListF X
| Cons : A -> X -> ListF X.

Definition inList : ListF List -> List := inn ListF.
Definition outList : List -> ListF List := out ListF (fold ListF).
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{verbatim}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists}
  \label{fig:listf}
\end{figure}

For purposes of presenting the interface for subsidiary recursion, let us consider three files in our codebase:

\begin{itemize}
\item \texttt{Subrec.v}: parametrized by a signature functor \verb|F| of type \verb|Set -> Set|,
  this provides (among much else, to be discussed below) \verb|Subrec : Set|, as the type to use
  for subsidiary recursion; and also a constructor \verb|inn : F Subrec -> Subrec| for that type.

\item \texttt{List.v}: parametrized by the type \verb|A : Set| of list
  elements, this file specializes the development in \verb|Subrec.v|
  to the case of lists.  This is done using defining the signature
  functor \verb|ListF|, shown in Figure~\ref{fig:listf}, giving us the
  type \verb|List|.  This is not to be confused with the type
  \verb|list| of lists in Coq's standard library.  The figure also
  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
  and types for conversion functions between \verb|List| and
  \verb|list| (code omitted).

\item \texttt{Span.v}: this implements the \verb|span| function
  mentioned above, as an \emph{algebra}.

\end{itemize}

\subsection{Algebras for subsidiary recursion}

Our approach is within a long line of work using ideas from
universal algebra and category theory to implement inductive datatypes
and recursion principles, in type theory.  On this approach, one describes
transformations to be performed on data as \emph{algebras}, which can then
be \textit{folded} over data.

\begin{figure}
\begin{verbatim}
Definition KAlg  : Type := Set -> (Set -> Set) -> Set.

Definition FoldT(alg : KAlg)(C : Set) : Set :=
  forall (X : Set -> Set) (FunX : Functor X), alg C X -> C -> X C.

Definition AlgF(Alg: KAlg)(C : Set)(X : Set -> Set) : Set :=
  forall (R : Set)
         (reveal : R -> C)        
         (fold : FoldT Alg R)
         (eval : R -> X R)      
         (d : F R),             
         X R.

Definition Alg : KAlg := MuAlg AlgF.
\end{verbatim}
\caption{The interface for algebras}
\label{fig:algf}
\end{figure}

The interface for algebras is presented in Figure~\ref{fig:algf}.
Let us consider the definitions.  \verb|KAlg| is the kind for the type-constructor for algebras,
as we see in the very last definition of the figure, for \verb|Alg|.  This type-constructor \verb|Alg|
is a fixed-point of the type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|,
to be discussed in Section~\ref{sec:mu} below.  Doing so requires that \verb|AlgF| only use its parameter
\verb|Alg| positively.  We will confirm this shortly.

The type \verb|FoldT Alg C| is the type for fold functions which apply
algebras of type \verb|Alg| to data of type \verb|C|.  At the top
level of code, \verb|C| would just be \verb|List| (for example).  When
one initiates a subsidiary recursion, though, this type \verb|C| will
instead by what we like to call the \emph{recursion universe} of that
recursion.  Each recursion is based on an abstract type \verb|R|,
representing the data upon which we will recurse.  Indeed, \verb|AlgF|
says that algebras take in such an \verb|R| as their first argument.
This means that an algebra can assume nothing about \verb|R| except
that it has the operations given next in the listing of \verb|AlgF|.
Thus, \verb|R| functions like a universe for recursion: one may
perform the various given operations, without leaving the type
\verb|R| (which is good, because we may recurse on elements of type \verb|R|).

And what are the operations on \verb|R|?  First there is \verb|reveal|, which
turns an \verb|R| into a \verb|C|.  This reveals that the data of type
\verb|R| are really lists (if this is a top-level recursion) or else
really belong to some outer recursion universe (if this is an inner
recursion).  Next we have \texttt{fold}, which will allow us to fold
another algebra over data of type \verb|R|.  We will use \verb|fold|
to initiate subsidiary recursions.  Then there is \verb|eval|, which
is used to make recursive calls on data of type \verb|R|.

This is a good place to highlight that for subsidiary recursion, it is
crucial that algebras have a carrier \verb|X| which depends
(functorially) on a type.  This is so that (i) inside an inner
recursion we may compute a result of some type that may mention
\verb|R|, but (ii) outside that recursion, the result will mention the
type \verb|C|.  The \verb|eval| function returns something of type
\verb|X R|, demonstrating (i).  For (ii): if we look at the definition
of \verb|FoldT| in the figure, we see that folding an algebra of type
\verb|alg C X| over a value of type \verb|C| produces a result of type
\verb|X C|.

\section{Examples of subsidiary recursion}
\label{sec:examples}

\section{Interface for subsidiary induction}
\label{sec:interfacei}

\section{Examples of subsidiary induction}
\label{sec:interfacei}

\section{Derivation of subsidiary recursion}
\label{sec:deriv}

\subsection{Retractive-positive recursive types}
\label{sec:mu}

\section{Related Work}
\label{sec:related}

In some tools, like Coq, Agda, and Lean, termination is checked
statically, based on structural decrease at recursive calls.  Others,
like Isabelle/HOL, allow one to write recursions first, and prove
(possibly with automated help) their termination
afterwards~\cite{krauss}.

It has not escaped the notice of designers of ITPs that structural
recursion is not the only form of terminating recursion.  All the
mentioned tools provide support for well-founded recursion, where for
recursive calls, one must show that the parameter of recursion has
decreased in some well-founded order.

Subsidiary recursion can be seen as a generalization
of nested recursion, which is a form of recursion allowing recursive
calls of the form \verb|f\ (f\ x)|.  In subsidiary recursion, these
are generalized to the form \verb|f\ (g\ x)|, where \verb|g| is
another recursively defined function.

\bibliography{biblio}
